---
title: "What Operational Factors Influence the Duration of Legal Matters, and What Process Improvements Can Reduce Resolution Time?"
author: "Elizabeth Ero"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    smooth-scroll: true
    code-fold: true
    css: styles.css
---

# 1. Executive Summary

This study investigates the operational factors that influence the duration of legal matters within an upstream oil and gas legal environment and evaluates how legal operations can improve turnaround efficiency. In fast-paced corporate environments, prolonged legal review timelines may delay commercial execution, increase transaction costs, and create operational bottlenecks for business units. The study therefore seeks to determine which operational variables most significantly contribute to extended legal matter duration.

The analysis is based on anonymised internal legal operations data collected from contract review trackers and email correspondences, covering the period from May 2021 to May 2026. The dataset contains 100 legal contracts across commercial and technical categories.

Using exploratory data analysis (EDA), data visualisation, hypothesis testing, correlation analysis, and regression modelling, the study examines the relationship between legal matter duration and operational variables such as matter complexity, revision cycles, approval layers, matter category, and counsel type.

Preliminary findings suggest that legal matter duration is strongly associated with operational complexity, the number of revision cycles, and approval escalation structures. The study concludes that streamlining approval processes, reducing unnecessary negotiation loops, and introducing standardised contract workflows may significantly reduce legal turnaround time and improve operational efficiency within the legal function.

# 2. Professional Disclosure

## Professional Background

I am a legal counsel in an upstream oil and gas company. My task involves legal advisory, contract drafting and negotiation, transaction support, corporate governance, and regulatory compliance activities within the oil and gas industry. My role requires management and review of a wide range of legal matters, including commercial agreements, regulatory filings, joint venture arrangements, procurement contracts, employment-related matters, and stakeholder negotiations.

A critical aspect of my responsibilities involves ensuring that legal matters are resolved efficiently without compromising legal quality, regulatory compliance, or commercial protection. Delays in legal review processes can affect transaction execution timelines, procurement activities, operational approvals, and broader business objectives. Consequently, understanding the factors that drive legal matter duration is operationally important to both the legal department and the wider business.

This study directly aligns with my day-to-day professional responsibilities because it evaluates how operational processes within legal departments influence turnaround time and identifies opportunities for workflow improvement and process optimisation.

## Operational Relevance of Techniques

### Exploratory Data Analysis (EDA)

Exploratory Data Analysis is operationally relevant because legal matter records often contain inconsistencies, missing values, outliers, and uneven distributions that may distort operational insights if left unaddressed. EDA enables the identification of bottlenecks, unusually delayed matters, and data quality issues that may affect legal operations reporting and decision-making. Within the context of this study, EDA assists in understanding how legal matters are distributed across duration ranges, matter types, and operational complexity levels.

### Data Visualisation

Data visualisation is particularly important within legal operations because senior management and business stakeholders often require simplified visual summaries of operational performance rather than technical statistical outputs. Visualisations such as boxplots, histograms, scatterplots, and correlation heatmaps help communicate trends in legal turnaround time, operational workload, and process inefficiencies in a manner that supports business decision-making. In this study, visualisation techniques are used to demonstrate how matter complexity, approval structures, and revision cycles influence legal matter duration.

### Hypothesis Testing

Hypothesis testing supports evidence-based legal operations management by determining whether observed differences in legal matter duration are statistically significant rather than random occurrences. For example, the study evaluates whether matters handled by external counsel differ significantly in duration from matters handled internally, and whether certain categories of legal matters systematically require longer resolution periods. These findings may support decisions regarding resource allocation, external counsel engagement, and workflow redesign.

### Correlation Analysis

Correlation analysis is operationally useful because legal processes are often influenced by multiple interconnected factors. Understanding the strength and direction of relationships between operational variables allows legal management to identify which factors are most strongly associated with delays. In this study, correlation analysis helps determine whether variables such as complexity score, revision count, and approval layers are positively associated with increased matter duration.

### Regression Analysis

Regression analysis is highly relevant to legal operations because it enables the simultaneous assessment of multiple operational variables affecting legal matter duration. Unlike simple descriptive analysis, regression modelling quantifies the independent impact of each predictor variable while controlling for other factors. This allows the legal department to identify which operational factors contribute most significantly to prolonged turnaround time and to prioritise process improvement initiatives accordingly. The findings from the regression model are expected to provide actionable recommendations for reducing legal matter resolution time and improving operational efficiency.

# 3. Data Collection and Sampling

## Data Source

The dataset used for this study was collected from anonymised internal legal operations records maintained within the legal function. The primary sources of data include:

- Legal department tasks tracker
- Email correspondences
- Approval workflow records

The dataset captures operational information relating to legal matters handled within the organisation between May 2021 and May 2026.

## Sampling Method and Data Collection

A purposive sampling approach was adopted to ensure that the dataset adequately represented the different categories of legal matters handled within the legal function. The sample includes commercial and technical matters.

To satisfy the minimum assessment requirements, the dataset contains 100 observations and includes both numeric and categorical operational variables relevant to legal workflow analysis. Each observation represents a distinct legal matter recorded within the legal operations process.

The selected time period was considered appropriate because it captures sufficient operational variation across multiple business cycles, transaction types, and organisational approval processes.

The data collection process involved extracting operational information from internal legal tracking systems and manually consolidating the records into a structured spreadsheet for analysis. Sensitive information such as contract titles, company names, personnel names, and commercially confidential information were excluded or anonymised before analysis.

Each legal matter was assigned a unique anonymised identifier, and operational variables relevant to the study objective were recorded, including:

Matter duration (days);
Matter type;
Counsel type;
Complexity score;
Revision count; and
Approval layers.

The final dataset was cleaned and standardised prior to analysis to improve consistency and reduce data quality issues.

## Variables Collected

| Variable | Type | Description |
|------------------------|------------------------|------------------------|
| Matter ID | Identifier | Unique anonymised matter reference |
| Duration Days | Numeric | Number of days between opening and closure |
| Complexity Score | Numeric | Complexity rating (1–5) |
| Revision Count | Numeric | Number of negotiation/review cycles |
| Approval Layers | Numeric | Number of approval stages |
| Counsel Type | Categorical | In-house or External |
| Matter Type | Categorical | Commercial or Technical |
| Open Date | Date | Date matter commenced |

## Ethical Considerations

Given the confidential nature of legal operations data, all personally identifiable information (PII), commercially sensitive information, and proprietary business information were removed before analysis. No counterparty identities, or internal personnel names are disclosed in this study. Generic upstream contract names were utilised to identified contracts.

The study is conducted strictly for academic purposes, and the dataset used in the analysis is anonymised to protect organisational confidentiality and comply with internal data privacy expectations.

# 4. Data Description

## Load Required Packages

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(corrplot)
library(effsize)
```

## Import Dataset

```{r}
legal_data <- read.csv("Data Analytics Exam - Legal Matters Data.csv")
```

## Data Cleaning

```{r}
# Clean column names
names(legal_data) <- gsub("\\.", "_", names(legal_data))

# Convert date variable and remove duplicates
legal_data <- legal_data %>%
  mutate(
    Open_Date = as.Date(Open_Date, format = "%d/%m/%Y"),
    
    # Standardise categorical variables (remove extra spaces)
    Counsel_Type = stringr::str_trim(Counsel_Type),
    Matter_Type  = stringr::str_trim(Matter_Type)
  ) %>%
  distinct()
```

## View Dataset

```{r}
head(legal_data)
```

## Dataset Structure

```{r}
str(legal_data)
```

## Summary Statistics

```{r}
summary(legal_data)
```

## Missing Values

```{r}
colSums(is.na(legal_data))
```

## Data Types

```{r}
sapply(legal_data, class)
```

## Data Quality Assessment and Cleaning

Several data quality checks were conducted prior to analysis to improve the reliability and consistency of the dataset.

### Outlier Detection

Exploratory analysis identified several unusually long-running legal matters, including observations with substantially higher duration values than the dataset average. These outliers were reviewed and retained because they reflected genuine operational realities rather than data-entry errors.

### Date Formatting Inconsistencies

The `Open.Date` variable contained inconsistent date formatting styles across records. Date values were standardised into a consistent date format to support accurate time-based analysis.

### Categorical Variable Standardisation

Categorical fields such as `Counsel.Type` and `Matter.Type` were reviewed for naming consistency and formatting uniformity to ensure accurate grouping and statistical testing.

### Missing Value Validation

A missing-value assessment confirmed that the final dataset contained no missing observations across the selected analytical variables.


# 5. Exploratory Data Analysis (EDA)

## Distribution of Matter Duration

```{r}
ggplot(legal_data, aes(x = `Duration_Days`)) +
  geom_histogram(bins = 20, fill = "steelblue") +
  labs(
    title = "Distribution of Legal Matter Duration",
    x = "Duration (Days)",
    y = "Frequency"
  )
```

### Interpretation

This distribution evaluates whether matter duration is normally distributed or skewed.

## Outlier Detection

```{r}
ggplot(legal_data, aes(y = `Duration_Days`)) +
  geom_boxplot(fill = "orange") +
  labs(
    title = "Outlier Detection for Matter Duration",
    y = "Duration (Days)"
  )
```

### Interpretation

This boxplot identifies unusually long-running legal matters.

## Duration by Counsel Type

```{r}
ggplot(legal_data, aes(x = `Counsel_Type`, y = `Duration_Days`)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Duration by Counsel Type",
    x = "Counsel Type",
    y = "Duration (Days)"
  )
```

### Interpretation

This visualisation compares matter duration between in-house and external counsel.

## Duration by Matter Type

```{r}
ggplot(legal_data, aes(x = `Matter_Type`, y = `Duration_Days`)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Duration by Matter Type",
    x = "Matter Type",
    y = "Duration (Days)"
  )
```

### Interpretation

This chart compares legal matter duration across different categories.

# 6. Data Visualisation

## Complexity Score vs Duration

```{r}
ggplot(legal_data, aes(x = `Complexity_Score`, y = `Duration_Days`)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Complexity Score vs Duration",
    x = "Complexity Score",
    y = "Duration (Days)"
  )
```

### Interpretation

This visualisation evaluates whether higher complexity increases legal matter duration.

## Revision Count vs Duration

```{r}
ggplot(legal_data, aes(x = `Revision_Count`, y = `Duration_Days`)) +
  geom_point(color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Revision Count vs Duration",
    x = "Revision Count",
    y = "Duration (Days)"
  )
```

### Interpretation

This chart examines whether additional negotiation cycles increase matter duration.

## Approval Layers vs Duration

```{r}
ggplot(legal_data, aes(x = `Approval_Layers`, y = `Duration_Days`)) +
  geom_point(color = "purple") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Approval Layers vs Duration",
    x = "Approval Layers",
    y = "Duration (Days)"
  )
```

### Interpretation

This visualisation assesses whether approval complexity affects turnaround time.

# 7. Hypothesis Testing

## Hypothesis 1

### Research Objective

To determine whether legal matter duration differs significantly between in-house and external counsel.

### Hypotheses

- H0: There is no significant difference in matter duration between counsel types.
- H1: There is a significant difference in matter duration between counsel types.

## t-Test

```{r}
t_test_result <- t.test(`Duration_Days` ~ `Counsel_Type`, data = legal_data)

t_test_result
```

## Effect Size

```{r}
cohen.d(`Duration_Days` ~ `Counsel_Type`, data = legal_data)
```

### Interpretation

The p-value and effect size determine whether counsel structure materially affects turnaround time.

## Hypothesis 2

### Research Objective

To determine whether matter duration differs significantly across matter categories.

### Hypotheses

- H0: Mean duration does not differ across matter types.
- H1: Mean duration differs across matter types.

## ANOVA

```{r}
anova_model <- aov(`Duration_Days` ~ `Matter_Type`, data = legal_data)

summary(anova_model)
```

### Interpretation

This analysis evaluates whether certain legal matter categories systematically require more time.

# 8. Correlation Analysis

## Select Numeric Variables

```{r}
numeric_data <- legal_data %>%
  select(`Duration_Days`,
         `Complexity_Score`,
         `Revision_Count`,
         `Approval_Layers`)
```

## Correlation Matrix

```{r}
cor_matrix <- cor(numeric_data, use = "complete.obs")

cor_matrix
```

## Correlation Heatmap

```{r}
corrplot(
  cor_matrix,
  method = "color",
  addCoef.col = "black"
)
```

### Interpretation

This analysis identifies the strongest operational relationships associated with legal matter duration.

# 9. Regression Analysis

## Regression Model

```{r}
regression_model <- lm(
  `Duration_Days` ~
    `Counsel_Type`+
    `Matter_Type` +
    `Complexity_Score` +
    `Revision_Count` +
    `Approval_Layers`,
  data = legal_data
)

summary(regression_model)
```

### Interpretation

This model estimates the combined effect of operational variables on matter duration.

## Regression Diagnostics

```{r}
par(mfrow = c(2, 2))

plot(regression_model)
```

### Interpretation

Diagnostic plots assess: - normality, - homoscedasticity, - linearity, - influential observations.

# 10. Integrated Findings and Recommendations

The analyses conducted in this study collectively demonstrate that the duration of legal matters is influenced not only by the inherent complexity of legal work, but also by operational workflow structures within the legal function. Across the exploratory analysis, visualisations, hypothesis testing, correlation analysis, and regression modelling, several operational variables consistently emerged as significant contributors to prolonged legal matter resolution time.

The exploratory data analysis revealed that legal matter duration is highly unevenly distributed. While a significant number of matters were resolved within relatively short timeframes, a smaller number of matters experienced substantially prolonged durations, creating a right-skewed distribution. This suggests that certain operational conditions or workflow characteristics create bottlenecks capable of significantly delaying matter completion.

The visualisation analysis further demonstrated that legal matters with higher complexity scores, greater revision counts, and more approval layers generally recorded longer turnaround times. In particular, the relationship between revision cycles and duration appeared consistently strong across multiple analyses, indicating that repeated negotiation and drafting cycles are a major operational driver of delay within legal workflows.

The hypothesis testing results provided statistical evidence that legal matter duration differs significantly across counsel structures and matter categories. Matters handled through external counsel recorded significantly longer average durations compared to internally managed matters. This finding suggests that coordination requirements, external review timelines, and communication inefficiencies may materially affect turnaround performance. Similarly, the ANOVA results indicate that certain categories of legal matters inherently require longer processing periods due to their operational and regulatory complexity.

The correlation analysis identified a particularly strong positive relationship between revision count and legal matter duration, suggesting that repeated negotiation rounds substantially increase completion timelines. Complexity score also demonstrated a meaningful positive relationship with duration, indicating that more sophisticated or operationally sensitive matters require longer review and approval periods.

The regression model provided the most comprehensive insight into the combined effects of operational variables on legal matter duration. The model explained approximately 77.6% of the variation in legal matter duration, indicating strong explanatory power. Most notably, revision count emerged as the strongest statistically significant predictor of prolonged duration. The findings suggest that each additional revision cycle contributes materially to increased turnaround time, even when controlling for counsel type, matter type, approval structures, and complexity levels.

Collectively, the analyses indicate that legal matter duration is not solely determined by legal complexity, but is heavily shaped by internal operational processes, negotiation structures, approval workflows, and coordination mechanisms. The findings therefore support the conclusion that meaningful improvements in legal turnaround time can be achieved through operational process optimisation rather than simply increasing legal staffing capacity.

**Recommendations**

Based on the findings of this study, several operational and process-improvement recommendations are proposed to improve legal matter turnaround time and enhance the efficiency of legal operations within the organisation.

---

## 1. Reduce Excessive Revision Cycles Through Standardisation

The analysis identified **revision count as the strongest predictor of prolonged legal matter duration**, indicating that repeated drafting and negotiation cycles significantly delay matter completion.

To address this, the legal function should expand the use of **standardised contract templates, pre-approved clauses, fallback language libraries, and negotiation playbooks**. For routine and low-risk matters, standard templates should be adopted more consistently to minimise unnecessary iterations and improve drafting efficiency.

Clearer internal negotiation parameters will also help reduce repeated business-side amendments and improve overall turnaround time without compromising legal quality or compliance.

---

## 2. Streamline Approval Structures

The findings suggest that **additional approval layers contribute to longer matter duration**, reflecting potential procedural inefficiencies.

While governance remains essential, the organisation should adopt a **risk-based approval framework**, where low-risk or routine matters follow simplified approval routes, while high-risk or strategic matters undergo enhanced review. Delegated approval authority for standard transactions may also reduce bottlenecks and improve responsiveness.

---

## 3. Improve Early Stakeholder Alignment

Frequent revisions may be driven by **incomplete instructions or shifting stakeholder expectations during drafting stages**.

To mitigate this, the organisation should strengthen **pre-engagement alignment processes**, ensuring clearer commercial instructions, defined negotiation objectives, and earlier engagement between legal and business teams before drafting begins. Structured kickoff meetings can help reduce ambiguity and improve efficiency.

---

## 4. Develop Complexity-Based Workflow Segmentation

The results show that **matter complexity influences legal turnaround time**, although not as strongly as revision cycles.

The legal department should classify matters based on complexity and risk, allowing for differentiated workflows:

- Low-complexity matters: streamlined, template-driven processes  
- High-complexity matters: enhanced review and dedicated oversight  

This approach will improve resource allocation and prevent routine matters from being unnecessarily delayed.

---

## 5. Strengthen Legal Operations Monitoring and Reporting

The study highlights the importance of operational visibility in managing legal performance.

The organisation should strengthen legal operations reporting through dashboards tracking key metrics such as:

- turnaround time  
- revision frequency  
- approval cycle duration  
- backlog levels  
- external counsel performance  

This will enable earlier identification of bottlenecks and support continuous process improvement.

---

## 6. Review External Counsel Engagement Processes

The findings indicate that **matters handled by external counsel take significantly longer than internal matters**.

To improve efficiency, the organisation should review external counsel management practices, including:

- clearer service-level expectations  
- standardised reporting timelines  
- improved coordination mechanisms  
- earlier alignment between internal and external teams  

Where appropriate, routine matters may be retained internally to reduce delays and improve control.

---

## 7. Introduce Legal Operations Analytics as a Continuous Improvement Tool

This study demonstrates the value of analytics in identifying inefficiencies in legal workflows.

The organisation should institutionalise **legal operations analytics** to support continuous improvement by tracking trends in:

- turnaround time  
- revision cycles  
- workload distribution  
- process bottlenecks  

Embedding data-driven decision-making into legal operations will enhance efficiency, transparency, and strategic support to the business.

---

**Overall, these recommendations are designed to reduce avoidable operational friction, improve consistency in legal processes, and enhance the speed and predictability of legal service delivery.**

# 11. Limitations & Further Work

This study is limited by its focus on a single upstream oil and gas legal environment, which may restrict generalisability to other industries. It also relies on historical operational data that may contain inconsistencies or incomplete records despite data cleaning. In addition, some relevant variables such as stakeholder responsiveness, transaction value, and urgency classification were not available for analysis. Finally, the results show statistical associations rather than causal relationships, meaning conclusions about cause and effect should be interpreted with caution.

Future research could expand the analysis to multiple organisations and industries to improve generalisability and benchmarking. It could also apply predictive analytics and machine learning to forecast legal matter duration and identify delay risks. Time-series analysis may help assess improvements over time following process changes. Additionally, qualitative research could provide deeper insights into behavioural and organisational drivers of delays. Finally, future studies may incorporate broader operational metrics such as service levels, stakeholder satisfaction, and workload distribution to strengthen legal operations analytics.

# References

Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R.* Lagos Business School / markanalytics.online. <https://markanalytics.online>

R Core Team. (2024). *R: A language and environment for statistical computing.* R Foundation for Statistical Computing. <https://www.r-project.org/>

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis.* Springer.

# Appendix: AI Usage Statement

Artificial intelligence tools, including ChatGPT, were used to assist with structuring the Quarto document and generating sample code templates. However, all analytical decisions, interpretations, and recommendations were independently developed by the author.
